## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <-5) {
  print(a)
} else {
  print("its more than 5")
}

```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5){
  print("small")
} else if (a > 5 & a <=10) {
  print("medium")
} else if (a > 10 | a == 7) {
  print("large")
} else {
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a/2 != 0) {
  if (a == 5) {
    print("a is 5")
  }
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  # paste is a concatenation function
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# change the first, third, and sixth elements to the value 6
for (i in c(1,3,6)){
  a_vector[i] <-6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row,col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10){
  # increase b by 2
  b <- b + 2
  # print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
  
}

# when we need to do stuff repeatedly or when we want to test stuff
```

An example:

```{r}
# start with a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

# then we want a function that converts from C to F:
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) +32
  
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
test_mat
# 1 is for rows, 2 is for columns
apply(test_mat, 1, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow <- 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    # Problem 1: Mumbling
    # My Solution:
    accum <- function(s){
      s_list <- unlist(strsplit(s, ""))
      s_accum <- ""
      counter <- 1
      while(counter <= length(s_list)){
        letter <- toupper(s_list[counter])
        s_accum <- paste(s_accum, letter, sep = '')
        if(counter > 1){
          sc <- 1
          while(sc<counter){
            letter <- tolower(letter)
            s_accum <- paste(s_accum, letter, sep = '')
            sc <- sc + 1
          }
        }
        if (counter != length(s_list)){
          s_accum <- paste(s_accum, '-', sep = '')
        }
        counter <- counter + 1
      }
      return(s_accum)
    }

    # Problem 2: Complementary DNA
    DNA_strand <- function(dna){
      #note to self the pairs are A-T and C-G
      dna_list <- unlist(strsplit(dna, "")[[1]])
      rna_list <-c()
      for (symbol in dna_list) {
        if(symbol =="A"){
          rna_list[length(rna_list)+1] <- "T"
        } else if(symbol == "T"){
          rna_list[length(rna_list)+1] <- "A"
        } else if(symbol == "C"){
          rna_list[length(rna_list)+1] <- "G"
        } else {
          rna_list[length(rna_list)+1] <- "C"
        }
      }
      return(paste(rna_list,collapse=''))
    }

    # Problem 3: Descending Order
    descending_order <- function(num){
      digits <- as.numeric(unlist(strsplit(as.character(num), "")[[1]]))
      digit <- sort(digits,decreasing=TRUE)
      final_num <- as.numeric(paste(digit,collapse=''))
      return(final_num)
    }

    # Problem 4: Convert a Boolean to a String
     boolean_to_string <- function(b){
       if (b){return("TRUE")}
       else{return("FALSE")}
     }

    # Problem 5: Highest and Lowest
    high_and_low <- function(numbers){
      num_list <- strsplit(numbers, " ")
      min_num <- as.numeric(num_list[[1]][[1]])
      max_num <- as.numeric(num_list[[1]][[1]])
      for(c_num in num_list[[1]]){
        num <- as.numeric(c_num)
        if (num>max_num){
          max_num <- num
        } else if (num < min_num){
          min_num <- num
        }
      }
      return(paste(max_num, min_num))
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    # map over a matrix of # for bonus points

    fizz_or_buzz <- function(num){
      soda <- "stale"
      
      if(num %% 3 == 0){
        soda <- "fizz"
        if(num %% 5 == 0){
          soda <- paste(soda, "buzz", sep = "")
        }
      }
      
      if(num %% 5 == 0){
        soda <- "buzz"
      }
      
      return(paste(num," is ",soda, sep = ""))
    }

    test_matrix <- matrix(c(3,4,5,6,7,8,9,10,11,12,13,14,15,16,17), nrow <- 5)
    for(row in 1:5){
      for(col in 1:3){
        print(fizz_or_buzz(test_matrix[row,col]))
      }
    }
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    marioStairs <- function(levels){
      for(level in 1:levels){
        steps <- ""
        for(step in 1:level){
          steps <- paste(steps, "X", sep = "")
        }
        print(steps)
      }
    }

    marioStairs(4)
    ```
